---
title: "Data frames and data import/export"
author: "Hugo and Nadia"
data: "BTM 2016"
output: html_document
---

```{r setup, echo=FALSE}
options(width = 110)
```

# Data frame

* Data frame -- 2-dimensional table with data (Excel spreadsheet, SQL-table)
* De facto it is standard way of data storage as "observations/variables": <br> rows are associated with observations, columns -- with variables

Places        N\_coord     E\_coord           Date     Temp
-------      ---------    ---------     ----------   ------
MM Faculty       59.88        29.83     07.01.2014      -15
PP Fortress      59.95        30.32     15.05.2013       17
SPbU main        59.94        30.3      22.06.2013       22
MM Faculty       59.88        29.83     09.01.2014      -21
...                ...          ...            ...      ...

Data frame inherited properties of matrices (rectangular shape) and lists (variables can have different types).

## Creating data frames

```{r}
LETTERS
```

```{r}
df <- data.frame(x = 1:4, y = LETTERS[1:4], z = c(T, F))
df
```

Function `str` gives information about the object and types of its variables:
```{r}
str(df)
```

## Names

```{r}
df <- data.frame(x = 1:4, y = LETTERS[1:4], z = c(T, F), 
                 row.names = c("Alpha", "Bravo", "Charlie", "Delta"))
df
```

```{r}
row.names(df)
colnames(df)
dimnames(df)
```

## Dimensions

```{r}
nrow(df); ncol(df) 
dim(df)
```

Important peculiarities:

* `length(df)` returns the number of _columns_ (variables), not the overall element number
* `names(df)` also returns only column names

```{r}
length(df); names(df)
```

## Data frame indexing

Indexing works the same way as for matrix:

```{r}
df[3:4, -1]
```

```{r}
df[, 1]; df[, 1, drop = FALSE]
```

Or for lists:

```{r}
df$z
df[[3]]
df[["z"]]
```

## Subsetting using a condition

You can subset of rows/columns of a data frame by either using `subset` function or directly specifying the condition you want to be fulfilled for your subset:

```{r}
df[df$x > 2, ]
```

```{r}
subset(df, x > 2)
```

```{r}
subset(df, x > 2, select = c(x, z))
```

## Combining data frames

Functions `rbind`/`cbind` work the same way as for matrices:

```{r}
rbind(df, data.frame(x = 5:6, y = c("K", "Z"), z = TRUE, row.names = c("Kappa", "Zulu")))
```

```{r}
cbind(df, data.frame(season = c("Summer", "Autumn", "Winter", "Spring"), temp = c(20, 5, -10, 5)))
```

## Combining data frames: `merge`

```{r}
df
```

Function `merge` allows you to merge data frames by a common variable:

```{r}
df_salary <- data.frame(x = c(3, 2, 6, 1), salary = c(100, 1000, 300, 500))
merge(df, df_salary, by = "x")
```

***

# Importing data

The easiest way to save and load data is using `".RData"` format:
```{r}
save(df, file="my_data_frame.RData")
load("my_data_frame.RData")
```
You can save multiple objects or even the whole environment in the `.RData` file. But it's gonna be heavy and not readable by anything else than R.

Usually the data is stored in tables or text/data files, which you will read with `read.*` function.

Main instrument -- `read.table` function:

* `file` -- name of a file
* `header` -- presence or absence of the header (names of columns) in the first row
* `sep` -- separator of values
* `dec` -- decimal point
* `quote` -- symbol denoting quotation marks for characters
* `na.strings` -- strings encoding missing values
* `colClasses` -- types of columns (to speed up reading or huge dataframes)
* `comment.char` -- symbol denoting comments
* `skip` -- number of rows to skip from the beginning of the file

Functions `read.csv`, `read.csv2`, `read.delim` и `read.delim2` are essentially spans over `read.table` function with different defaults for parameters.

## Typical pre-processing workflow:

* Importing data into a data frame
* Cleaning the values, checking the types
* Working with strings: names, character variables, factors
* Missing values: indentification, processing
* Manipulating variables: transformation, creation, removal
* Descriptive statistics calculation: split-apply-combine
* Visualisation
* Export

## Cleaning the data, checking the types

* Numeric types can become characters:
    + because of missing values designated with anything else than `NA`
      `na.strings = c("NA", "Not Available", "Missing")`
    + because of wrong separators or decimal signs <br>
      `sep = ","`, `dec = "."`
    + because of wrong quotation marks, supplementary text or comments <br>
      `quote`, `comment.char`, `skip`
>- Characters can become factors and the other way round <br>
   `as.character`, `as.factor`

Very useful functions:
* `str` -- check types of the variables
* `summary` -- check summary statistics (mean, median, min and max)
* `head` -- first 6 rows of a data frame
* `tail`-- last 6 rows of a data frame

## Processing variables

* Functions `complete.cases` and `na.omit` to get rid of observations with missing values: <br>
   `df[complete.cases(df), ]` or `na.omit(df)`
* It's dangerous to replace NA with other values!
* Creating, altering and deleting variables <br>
    `df$new_var <- <...>`, <br>
    `df$old_var <- f(df$old_var)`, <br>
    `df$old_var <- NULL`

## Export

* `write.table`, `write.csv` and `write.csv2` -- almost identical to import functions
* If an array is big enough, it;s better to separate the pre-processing step
    + in a separate ".R" file
    + write the preprocessed ("clean") data in a separate file

## Playing with the dataset tutorial

Please visit [this webpage](http://tinyurl.com/BTM2016-RDatasets) to go on with analysing datasets.

## Hands-on part

**Exercise.** The built-in data set `mtcars` contains information about cars from a 1974 Motor Trend issue. 
Load the data set (`data(mtcars)`) and try to answer the following:

1. What are the variable names? (Try `names`.)
2. What is the maximum `mpg`? (Try `max`.)
3. Which car has this? (Try `which`.)
4. What are the first 5 cars listed?
5. What are all the values for the Mercedes 450slc (`Merc 450SLC`)? Save it in the file "Merc.dat" with `write.table`.
6. Make a scatterplot of cylinders (`cyl`) vs. miles per gallon (`mpg`). Fit a regression line. Is this a good candidate for linear regression?

## Solution

1.  What are the variable names?
```{r}
data(mtcars) # upload data
names(mtcars)
```
2. What is the maximum `mpg`?
```{r}
max(mtcars$mpg)
```
3. Which car has this?
```{r}
row.names(mtcars)[which(mtcars$mpg==33.9)]
# following strings will do the same:
row.names(mtcars)[which.max(mtcars$mpg)]
row.names(mtcars)[which(mtcars$mpg==max(mtcars$mpg))]
```
4. What are the first 5 cars listed?
```{r}
row.names(mtcars)[1:5]
```
5. What are all the values for the Mercedes 450slc (`Merc 450SLC`)? Save it in the file "Merc.dat" with `write.table`.
```{r}
mtcars[row.names(mtcars)=="Merc 450SLC",]
# Since row.names should be unique, this will do the same:
mtcars["Merc 450SLC",]
data_to_save <- mtcars[row.names(mtcars)=="Merc 450SLC",]
write.table(data_to_save, file="Merc.dat")
```
6. Make a scatterplot of cylinders (`cyl`) vs. miles per gallon (`mpg`). Fit a regression line. Is this a good candidate for linear regression?
```{r}
cor(mtcars$cyl,mtcars$mpg)
fit <- lm(mtcars$cyl ~ mtcars$mpg)
plot(mtcars$mpg,mtcars$cyl)
abline(fit,col="red")
```

# Acknowledgements

Huge thanks to [Anton Antonov](http://github.com/tonytonov) for his lectures on R on [stepik.org](http://stepik.org).