---
title: "Basic R course"
author: "Nadezda Volkova, Daniel Elias Martin Herranz, Jack Monahan"
date: "BTM 2016"
output: html_document
---
  
```{r setup, echo=FALSE}
options(width = 110)
```

# 1. Functions.

## What to do if nothing works

>- Calm down, it is fine
>- Read thoroughly all the error and warnings messages
>- Check out help page: `help("sin")`, `help.search("logarithm")`
>- R-FAQ: https://cran.r-project.org/doc/FAQ/R-FAQ.html
>- Use Google!
>- R Mailing List Archive
>- stackoverflow.com

  
## Global environment
  
* Every variable exists in its _environment_
* Variable's environment - a space where you can find it
* All the variable you create in console go to _global environment_
* You can see these variables if you click "Environment" in RStudio

```{r}
ls()
```

## Functions

R has a lot of built-in functions, i.e. `rnorm` (which returns a given number of values of a normally distributed random variable):

```{r}
?rnorm
## rnorm generates random deviates of the normal distribution:
## rnorm(n, mean = 0, sd = 1)
```

```{r}
rnorm(15)
```

You can list arguments by names:

```{r}
rnorm(15, mean = 5, sd = 3)
```

## Writing functions

Main word - `function`: then _arguments_ in parentheses, and _body_ in curved brackets.

```{r}
returntwo <- function() { 
  y <- 2
  return(y)
}

returntwo()
```

## Function environment

If a variable exists only within a function - you can not call it from outside:

```{r}
addten <- function(x) { 
  x <- x + 10
}

addten(1)
ls()
```

A function can alter the variable it is applied to:

```{r}
addten <- function(x) { 
  x <- x + 10
}

cats <- 1
addten(cats)
cats # notice that only this string goes to the output
```

Or just return some transformation of this variable:

```{r}
addten <- function(x) { 
  return(x + 10)
}

morecats <- addten(cats)
print(c(cats,morecats))
```

# Function environment

There is also a global assignment operator `<<-` in case you really need to alter some global variable inside a function.

```{r}
addten <- function(x) { 
  moredogs <<- x + 10
}

dogs <- 1
addten(dogs)
dogs
moredogs
```

* This is called a function with a _side effect_
* But that's not a very good thing to do

## Removing variables from environment

Check what variables do you have:

```{r}
ls()
```

And simply remove the ones you do not need:

```{r}
rm(dogs)
rm(moredogs)
ls()
```

Or delete everything!

```{r}
rm(list = ls())
ls()
```

But be careful when removing variables!


# 2. Packages


* R was created to be extendable - it is based on _packages_
* The core of the language -- `base` package
* There are several other packages uploaded with `base` by default, others can be called without installation
* Don't mix up `package` and `library`!
* Again: extension modules in R are called packages.
* Library (in R ecosystem) - place on the disk where all available packages are stored.

Check out where your packages are stored:

```{r}
.libPaths()
```

## Where to get packages

>- CRAN: https://cran.r-project.org/
>- github, bitbucket

With this command you can check what packages you have already installed:

```{r, eval=FALSE}
installed.packages()
```

```{r, echo = FALSE}
i <- as.data.frame(installed.packages())
i <- na.omit(i[, c("Package", "Priority")])
i[i$Priority == "base", ]
```

## Loading packages

>- `library(pkgname)`: write package name without quotes
>- Instead of `library(pkgname)` you can use `require(pkgname)`

## Installing packages from CRAN

This is how you install any package from CRAN, in this case "xts" package:

```{r, eval=FALSE}
install.packages("xts", dependencies = TRUE)
```

If you call `library()` for a package you did not install, you will get an error.

```{r, message=F, warning=F, eval=F}
library(xts)
```

It is nice to update packages from time to time: `update.packages()`

Another very useful function if something is not working as planned: `sessionInfo()`

```{r, echo=FALSE}
sessionInfo()
```

## Bonus: installation from BioConductor

As you may want to install stuff from [http://bioconductor.org]() as well, use this code:

```{r, eval=FALSE}
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite( <package you want to istall> )
```



# 3. Vectors, lists and matrices

## What is a vector
* Vector - basic unit of data structure in R
* More complex structures such as matrix, data frame, factor and lists are based on vectors
* Remember the word 'vectorisation' - that's the main concept

. . .

$$   
\begin{align}
  f: 
  \begin{bmatrix}
    x_{1} \\
    x_{2} \\
    \vdots \\
    x_{m}
  \end{bmatrix} 
  \longrightarrow
  \begin{bmatrix}
    f(x_{1}) \\
    f(x_{2}) \\
    \vdots \\
    f(x_{m})
  \end{bmatrix} 
\end{align}
$$

## Vector

* Vector is an indexed set of objects of the same type
* R does not differentiate scalar (of length 1) and vector objects
* Vector element enumeration starts with 1 (not 0!)
* To create a vector, you can use a function `vector`

```{r}
x <- vector(length = 2)
x[1] <- 5
x[2] <- 8
x
```

## Vector making: function `c`

Instead of elementwise assignment, you can do it with a more compact notation:
```{r}
x <- c(5, 8)
x
```

Function `c` can be nestes and take variables:
```{r}
y <- c(x, 1, c(3, 4), x, NA)
y
```

## Vector making: operator `:`

If you need to make a sequence of number with step 1:
```{r}
5:9
3:-1
```

## Vector making: function `seq`

Used to make more sophisticated sequences:
```{r}
seq(1, 2, by = 0.25)
```

```{r}
seq(3, 4, length.out = 5)
```

```{r}
seq(3, 4, length = 5)
```

## Vector making: function  `rep`

If you want to make a vector by repeating something:
```{r}
rep(1:3, times = 3)
```

```{r}
rep(1:3, each = 4)
```

```{r}
rep(1:3, length.out = 5)
```

## Types of vectors
* Atomic vectors: all elements are of the same type
* Main types:
  1. logical (TRUE/FALSE)
  2. integer
  3. numeric/double 
  4. complex
  5. character
  6. raw

## Types of vectors

If you would like to know what types does a vector have, use functions `typeof` and `is.*` (where `*` means any particular type)

```{r}
a <- c("Tree", "Flower", "Bird")
typeof(a)
```

```{r}
is.character(a)
is.logical(a)
```

## Type convertion

Use `as.*` to force a vector to turn into another type:
```{r}
b <- "123.25"
as.numeric(b)
```

```{r}
as.integer(b)
```

## Length of a vector: `length`

Every vector has a length:
```{r}
x <- 1:100
length(x)
```

It is determined automatically, but you can change it:
```{r}
length(x) <- 4; x
```

```{r}
length(x) <- 7; x
```

## Named vectors

Vector elements can be named:
```{r}
a <- c(uno = 1, dos = 2, "universal answer" = 42, 99)
names(a)
```

```{r}
names(a) <- c("one", "two", "forty two", "ninety nine")
a
```

```{r}
#names(a) <- NULL
```

## Vector arithmetics

Arithmetic operators are _vectorized_ by default (are applied elementwise):
```{r}
1:3 + c(-1, 2, 0)
```

```{r}
1:3 * c(-1, 2, 0)
```

```{r}
c(TRUE, TRUE, TRUE) & c(0, 1, 999)
```

## Vectorization

Many functions in R have a built-in vectorization support:
```{r}
# vectorized
sqrt(1:4)
```

```{r}
# vectorized
floor(seq(0, 3, by = 0.25))
```

```{r}
# not vectorized
sum(1:100)
```

## Recycling

How does the arithmetics work on vectors of different length?

1. Length of the results is equal to the length of the bigger vector

2. Smaller vector is recycled several times so that the length of overwritten vector equals the length of the bigger vector.

3. If the length of the bigger vector is not divisible by the length of smaller, you will get a warning

```{r, eval = FALSE}
1:5 + 0:1
# recycling in action: (1, 2, 3, 4, 5) + (0, 1, 0, 1, 0) => (1, 3, 3, 5, 5)
```

```{r, echo = FALSE}
1:5 + 0:1
```

Why do we need it?

```{r}
1:10 + 3
```

```{r}
(5:8) ^ 2 
```

```{r}
1:4 >= 3
```

## Accessing vector elements

Let's take a simple vector:
```{r}
x <- seq(10, 100, by = 10)
```

Reading and writing of vector elements is done by using the operator `[`:
```{r, eval = FALSE}
x[ind]
val <- x[ind]
x[ind] <- val
```

Trivial case:
```{r}
x[]
```

## Positive indices

```{r}
x[1]
```

```{r}
x[3:4]
```

```{r}
x[c(8, 7, 3, 6:8, x[1])]
```

## Negative indices

It means "All elements except for these":

```{r}
x[-5]
```

```{r}
x[-(2:6)]
```

```{r}
x[c(-3, -5, -length(x), -5)]
```

## Logical indices

Equals "Elements whose corresponding value is TRUE":

```{r}
x[rep(c(TRUE, FALSE), 5)]
```

```{r}
x[c(TRUE, FALSE)]
```

```{r}
x[x > 77 & x < 99]
```

## Indexing via names

For names vectors, you can use the dollowing construct:

```{r}
a[c("two", "one", "forty two")]
```

```{r}
a[c("forty two", "forty three", "forty four")]
```

## Functions `all` and `any`

```{r}
all(x < 200); all(x > 20)
```

```{r}
any(x > 150); any(x < 15)
```

## Function `which`

```{r}
which(x >= 50)
```

```{r}
which.min(x)
which.max(x)
```

## Matrix

* Matrix is a 2-dimensional array of data of the same type
* There is a function `matrix`

```{r}
matrix(1:6, nrow = 2, ncol = 3)
```

## Matrix making

You can shorten the notation without loosing any information:

```{r}
matrix(1:6, nrow = 2)
```

```{r}
matrix(1:6, nrow = 2, byrow = TRUE)
```

```{r}
matrix(7:8, nrow = 2, ncol = 5)
```

## Attribute `dim`

With this attribute (feature of an object) you can assess the dimensionality of your matrix:

```{r}
m <- matrix(1:6, ncol = 3)
dim(m)
c(nrow(m), ncol(m))
```

```{r}
dim(m) <- NULL; m 
dim(m) <- c(2, 3); m
```

## Arithmetic operators

All arithmetic operations process matrices elementwise:

```{r}
m1 <- matrix(1:4, nrow = 2) 
m2 <- matrix(c(1, 2, 2, 3), nrow = 2)
```

```{r}
m1 + m2
m1 + 5
```

```{r}
m1 * 2
m1 * m2
```

Linear algebra matrix multiplication should be written in `% %`:
```{r}
m1 %*% m2
```

## Indexing of matrices

Rules are the same as for vectors, but now we have 2 dimensions:

```{r}
m <- matrix(1:10, ncol = 5)
m[1, 3]
```

```{r}
m[2, ]
m[, 4]
```

```{r}
m[1, ] <- 0; m
```

```{r}
m[, -5] <- 11:18; m
```

## Collapsing dimensions

If you want to access a particular part of a matrix, you just treat it as a vector of rows and vector of columns:

```{r}
m <- matrix(1:10, ncol = 5)
ind <- c(1, 3, 5)
m[, ind]
```

```{r}
ind <- 3
m[, ind]
```

```{r}
m[, ind, drop = FALSE]
```

## Named matrices: `rownames`/`colnames`

If columns and rows of a matrix are named, they can also be called using these names:

```{r}
m <- matrix(1:10, ncol = 5)
rownames(m) <- c("row1", "row2")
colnames(m) <- paste0("column", 1:5)
```

```{r}
m
```

```{r}
m["row1", c("column2", "column4"), drop = F]
```

## Joining matrices: `rbind`/`cbind`

You can put two matrices together by joining their columns (`cbind`) or rows (`rbind`):

```{r}
rbind(m1, m2)
```

```{r}
cbind(m1, m2)
```

```{r}
cbind(m1, m2, 1:2, c(5, 3), m2[, 1], m1 * 3, cbind(m2, m1))
```

## `apply`

R has a very convenient way of applying functions to matrices. Let's define a matrix and a function we would like to apply.

```{r}
m <- matrix(1:25, 5)
f <- function(x) sum(x^2)
```

`apply` has 3 arguments: 

>- Array (matrix)
>- Index (1 -- row-wise, 2 -- column-wise)
>- Function

```{r}
apply(m, 1, f)
apply(m, 2, f)
```

```{r}
apply(m, 1:2, function(i) if (i>13) i else 13)
```

```{r}
m[m <= 13] <- 13; m
```

Such a function without name is called _anonymous_.

## `rowSums`, `rowMeans`, `colSums`, `colMeans`

The most common row-wise and column-wise operations are `sum` и `mean`:
```{r}
m <- matrix(1:25, 5)
rowSums(m)
```

Let's doublecheck that they actually work this way:
```{r}
all.equal(rowSums(m), apply(m, 1, sum))
all.equal(colMeans(m), apply(m, 2, mean))
```

## Lists

>- List is an indexed structure
>- List can consist of arbitrary objects
>- Usually used to store data of different length or different types
>- Function `list` creates a list

```{r}
list(1:5, "my_data", matrix(0, 2, 2))
```

```{r}
list(a = 1, b = 1:3, "1to5" = 1:5, 42)
```

A list can be recursive:

```{r, eval = FALSE}
list(a = list(1, 2, 3), b = list(list(4), 5, 6))
```

You can concatenate multiple lists into a new one:

```{r}
l1 <- list(name = "john", salary = 1000)
l2 <- list(has_car = TRUE, car = "lamborghini")
```

```{r}
c(l1, l2)
```

Every vector can be easily converted into a list:

```{r}
v <- 1:7
list(v)
```

But not the other way round! In cases when it makes sense, you can use `unlist` to make a vector out of list:

```{r}
l <- list(1:3, 4:5, last = 6)
unlist(l)
```

```{r}
unlist(c(l, "spy!"))
```

There are several ways to access elements of a list.

First method: as for vectors, `[]`
    + indexing according to vector rules
    + result is a sublist
    
```{r}
l[3:2]; l[-(1:2)]
```

```{r}
l[c(F, T, F)]; l["last"]
```

Second method: accessing a particular element, `[[]]`
    + (scalar) number of element or its full name
    + result is an element of the list

```{r}
l[[1]]
l[["last"]]
```

Third method: calling an element of a names list using its name, `$`
    + enough to use partial name of an element
    + results is an element of the list
    
```{r}
l$last
```

```{r}
l$l #l$la; l$las
```
    
## Replacing and adding elements to the list

```{r}
l <- list(1:3, 4:5, last = 6)
l[[3]] <- NULL; l
```

```{r}
l[[4]] <- 99; l
```

```{r}
l <- list(vec = 1:7, fun = sqrt) 
names(l)
is.null(l$string)
```

```{r}
l$string <- "Citius, altius, fortius"
l
```
    
    
## Applying functions to a list: `lapply` 

R also has a special function to apply functions to all elements of a list. Result will also be a list.

```{r}
l <- list(a = c("12", "34"), b = LETTERS[5:10], c = 1:5)
```

```{r}
lapply(l, length)
```

```{r}
lapply(l, paste, collapse = "|")
#lapply(l, function(s) paste(s, collapse = "|"))
```

There is also a function `sapply` that can be applied to a list or vector and returns a vector (when possible).

```{r}
sapply(l, paste, collapse = "|")
```



# 4. `if` и `else`

Decision making is an important part of programming. This can be achieved in R programming using the conditional __if...else__ statement.

```{r, eval=FALSE}
if (<condition>) {<do something>} else {<do another thing>}
```

* Here \<condition\> is an expression that results in logical vector of length 1 (`TRUE` or `FALSE`, `NA` will produce an error)
* \<do something\> and \<do another thing\> -- arbitrary commands
* You can put multiple conditions by using `|` as _OR_ or `&` as _AND_.

Check out this:

```{r}
if (sqrt(2) > 1.5) {
  print("Greater!")
} else {
  print("Less!")
}
```

**NB**: do not transfer `else` on a new string!

The above conditional can also be written in a single line as follows.

```{r, echo=FALSE}
if (sqrt(2) > 1.5) print("Greater!") else print("Less!")
```

## Nested __if...else__ statement

We can nest as many __if...else__ statement as we want as follows.

Syntax of nested if...else statement:

```{r, eval=FALSE}
if ( test_expression1) {
   statement1
} else if ( test_expression2) {
   statement2
} else if ( test_expression3) {
   statement3
} else
   statement4
```

Only one statement will get executed depending upon the test_expressions.

```{r}
x <- 0
if (x < 0) {
   print("Negative number")
} else if (x > 0) {
   print("Positive number")
} else
   print("Zero")
```

## `ifelse`

If you want to avoid all the hassle with if and else and calling different variables, there is a solution!

```{r}
if (c(TRUE, FALSE)) print("Hmm...?")
```

__ifelse__ returns a value with the same shape as test which is filled with elements selected from either yes or no depending on whether the element of test is TRUE or FALSE.

```{r}
ifelse(runif(8) > 0.5, "Heads", "Tails")
```

In other words, you write `ifelse( <yes/no> , <yes-action>, <no-action>)` if you need a response of the same length as condition. You can also put as many `ifelse` inside each other as you want.

```{r}
x <- runif(8)
ifelse(x > 2/3, "Rock", 
       ifelse(x > 1/3, "Scissors", "Paper"))
```

## Multiple choice: `switch`

There is a function __switch(EXPR, alternatives)__ that evaluates EXPR and respectively chooses one of further arguments as a response.

```{r}
switch("factorial", 
       sum = 5 + 5,
       product = 5 * 5,
       factorial = factorial(5), 
       0)
```

There is not so many cases where you could use it, but it's nice to know just in case.

## Loops

### `repeat`

When to use: when you want to run a loop forever and break it on your own will:

```{r}
i <- 0
repeat {
  i <- i + runif(1)
  print(i)
  if (i > 5) break
}
```

### `while`

When to use: when you want something to be happening as long as some condition is satisfied

```{r}
i <- 2^14
while (i > 1000) {
  i <- i/2
  print(i)
}
```

### `for`

When to use: when you are using some index that runs in pre-determined interval

```{r}
for (i in 1:8) {
  if (i %% 2 == 0) print(i)
}
```

Index is not necessarily integer:

```{r}
for (i in letters) {
  if (i == "b") next
  if (i == "d") break
  print(i)
}
```

**NB**: If you want to print something inside a loop, you have to call function `print`! Otherwise nothing happens:

```{r}
for (i in 1:5) i
```

### `for` vs vectorization

Somethimes running things in for-loop is time-comsuming (at least a little bit).

```{r, cache=TRUE}
v <- 1:1e5
system.time({
  x <- 0
  for (i in v) x[i] <- sqrt(v[i])
})
```

So why not using R's ability to perform vector calculations?

```{r}
system.time({
  y <- sqrt(v)
})
identical(x, y)
```


# 5. Strings

* String is an element of a vector of type `character`
* Most of the functions operating with strings are vectorized

```{r}
s <- c("Trump and republicans!", 
       "Clinton and democrates?",
       'Against all.')
```

```{r, echo=FALSE}
#s <- enc2utf8(s)
```

* Usually you would use quotation marks to designate a string
* But ' ' also work: `'"Smart" watch'`

```{r, echo=FALSE}
'"Smart" watch'
```

## Functions `paste`/`paste0`

* This functions perform concatenation of string vectors with respect to rewriting rules
* Argument `sep` -- separator between elements (by default, for `paste` it is space, for `paste0` -- empty line)
* Argument `collapse` -- "collapses" vector into one string

```{r}
paste(c("Beautiful", "Dirty"), "Rich")
```

```{r}
paste(c("Beautiful", "Dirty"), "Rich", sep = "_")
```

```{r}
paste(c("Beautiful", "Dirty"), "Rich", collapse = ", and")
```

## Function `strsplit`

* Splits a string vector
* **NB**: The result is a list!

```{r}
strsplit(s, " и ", fixed = TRUE)
```


If you do not put the argument `fixed` (by default `fixed = FALSE`), the string which you want to split will be considered a regular expression:

```{r}
strsplit(s, "[[:punct:]]")
```

## Regular expressions

* "Metalanguage" of search and substring manipulation
* The search template can include usual symbols and so-called _wildcards_

```{r}
grep("Trump", s)
```

```{r}
grepl("Trump", s)
```

```{r}
gsub("\\b[[:alpha:]]{4,5}\\b", "####",  s)
```

## Package `stringr`

Installation:
```{r}
#install.packages("stringr")
library(stringr)
```

This package contains a plenty of functions for string processing. All these functions are vectorized and consistent:

```{r}
str_extract(s, "")
```

```{r}
str_replace(s, "[all]", "?")
```

You can add `_all` to most of the functions to find all the occurences and not only the first one:

```{r}
str_extract_all(s, "m")
```

```{r}
str_replace_all(s, "[all]", "?")
```

## Functions `tolower`/`toupper`

Case manipulation:
```{r}
tolower(month.name)
```

```{r}
toupper(month.abb)
```

## Filepaths

An R session always has a _working directory_:
```{r}
getwd() 
head(list.files())
list.dirs("..", recursive = FALSE)
```

* Use the function `setwd` to change working directory.
* When you create a new project in RStudio, the working directory is the folder where project file `.Rproj` is located.
* Catalogues separators in Windows: either `/`, or `\\`

## Number formatting

```{r}
c(pi, exp(pi))
```

```{r}
formatC(c(pi, exp(pi)), digits = 3)
```

```{r}
formatC(c(pi, exp(pi)), digits = 3, format = "e")
```

## Function `cat`

* Usually the function `print` is used for output
* Function `cat` can print the objects "as they are"

```{r}
print('"Smart" watch'); cat('"Smart" watch')
```

```{r}
print("Harry\tRon\nHermione"); cat("Harry\tRon\nHermione")
```


# Tasks

**Excercise 1**. Create a function that takes argument n and calculates n-th element of [Fibonacchi](https://en.wikipedia.org/wiki/Fibonacci_number) sequence.

**Excercise 2**. Create a `k` x `k` matrix of zeros, put `1`s in the positions where $i+j = k+2$ and calculate the sum in every column.

**Excercise 3**. Imagine you have a list of observations of bacterial counts in different timepoints in several dishes:
```{r} 
bactcounts <- list(plate_1 = c(130,240,344,560), plate_2 = c(200,267,300,222), plate_3 = c(400,370,230,190))
```
Calculate the means and medians of values in every plate and create a named vector `MeMe` which will contain "yes" if a mean count for a plate is less than its median and "no" otherwise.


# Acknowledgements

Huge thanks to [Anton Antonov](http://github.com/tonytonov) for his lectures on R on [stepik.org](http://stepik.org).
