---
title: "Excercise 2"
author: "Your name"
date: "20 October 2016"
output: html_document
---

### Plots

Here is a simple scatter plot
```{r}
plot(1:10, main="My scatterplot")
```

Here are two plots side-by-side
```{r}
# Create plotting region with one row, two columns
par(mfrow=c(1,2))

# First plot
plot(1:10, main="My scatterplot")

# Second plot
hist(rnorm(100))
```

You can add a name for your chunk after the "r" symbol like this. Note: you shouldn't have two chunks with the same name!
```{r chunk_name}
paste("Hello world")
```


You can set figure widths and heights, either together or separately. 
```{r fig.width=5, fig.height=6}
plot(1:10, main="My scatterplot")
```

You can also align the figure to the `centre`, `left` or `right`
```{r fig.width=3, fig.align="right"}
plot(1:10, main="My scatterplot")
```

By default the resolution of your produced image is `75`. You can change this using the `dpi` option. See the difference between both plots below. 
```{r dpi=10, out.width="700px"}
plot(1:10, main="My scatterplot")
```

```{r dpi=150, out.width="700px"}
plot(1:10, main="My scatterplot")
```

Figures aren't saved to file by default. To save figures in a chunk to file set the `fig.path` to the directory to save the path. 

```{r figure_chunk, fig.path='chunk_figures/'}
plot(1:10, main="My scatterplot")
barplot(1:10, main="My barplot")
```

If you set this to `./` the figures will be saved to the directory containing the `.Rnw` file.

```{r figure_chunk_2, fig.path='./'}
plot(1:10, main="My scatterplot")
```


## Caching

Here is a cached chunk. The first time you run it, it will take ~5 seconds to generate because we are setting the system to sleep for 5 seconds before printing (we are mimicing a computationally intensive function). Any subsequent runs should not take as long, becase the results are cached on your machine. 

Play around with it a little. Try setting `cache=FALSE`. Does it run slower?
```{r, cache=TRUE}
# Sleep for five second
Sys.sleep(5)
# Print a result
paste("This took me a long time to run!")
```
